Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat

### 2023-10-19 manuscript.Rmd
some updates

---
title: "It's easier to move sideways"
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
        - nerr
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago. We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island. A total of 2284 individual crabs representing 90 morphospecies (19 families) were found. Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.  We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago. Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
bibliography: references/citations.bib
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: Table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

# Introduction {-}
Natural communities are structured by a suite of biotic and abiotic factors including geography, habitat complexity, environment, productivity, and species interactions (among many others) [@begonEcologyIndividualsEcosystems2021] and such complexity makes direct comparisons over large geographic scales challenging. Marine communities in particular experience unique influences and present research challenges not found in terrestrial habitats [@steeleComparisonTerrestrialMarine1985; @kinlanPropaguleDispersalMarine2003; @raffaelliMarineTerrestrialEcologists2005; @grosbergBiodiversityWaterLand2012; @knappDriversBiodiversityChange2017]. Much of our understanding of biotic zonation and community structure in the ocean comes from classic studies in the intertidal (some of the best-studied marine habitats) [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974]. Such studies point to a combination of physical and biological factors, including physiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities. Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999], a generalization has emerged for the intertidal that lower limits of species distributions are set by ecological interactions whereas upper limits are determined by abiotic factors.

@@

A<!--nother--> likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023a; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->

Studies examining community structure along depth or latitudinal gradients have been criticized for confounding multiple potentially covarying aspects of the environment (such as habitat area or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Here, we employed autonomous reef monitoring structures (ARMS) [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009], collection devices designed to mimic reef structural complexity and enable sampling of cryptic fauna in a standardized way. Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities. The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI) are the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) are the uninhabited atolls and banks northwest of Kaua&#x02BB;i. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to affect the variation in structure of sampled communities. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible drivers on community variation. Given the possibility for considerable separation between deep and shallow reefs [e.g., @hobanPlumbingDepthsEnvironmental2023a; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative effects of depth (shallow vs. mesophotic) and island region.   Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI). We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts). We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.  Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.

@@

```{r site-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

crab_sites <- cc$crabs_untransformed %>%
  sample_tibble() %>%
  distinct(hawaiian_name,island,island_group,lat,lon,.keep_all = FALSE) %>%
  mutate(
    island = case_when(
      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
      TRUE ~ as.character(hawaiian_name)
    ),
    island = fct_reorder(island,-lat)
  ) %>%
  # mutate( island = fct_reorder(island,-lat) ) %>%
  st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%

@@

box_before <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")))

@@

pmnm <- st_read(here("data","gis","pmnm.shp"),quiet = TRUE) %>%
  st_transform(transformer)

ggplot() +
  # geom_contour(data=bath_proj,mapping=aes(x=V1,y=V2,z = V3),  color="#cecece") +
  # geom_raster(data=bpr,aes(x=x,y=y,fill=z)) +
  # geom_tile(data=bath_proj,mapping=aes(x=V1,y=V2,fill = V3)) +
  # scale_fill_gradient2(low="dodgerblue4", mid="gainsboro", high="darkgreen") +
  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
  geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=island),size=3,shape=21,color="black") +
  scale_fill_manual(values=island_pal,name="Island") +
  coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
  scale_x_continuous(breaks = -seq(180, 155, -5)) +
  theme_minimal() +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
cat("{#fig:fig_map}")
```

@@

The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]). `r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated. The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).

```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

@@

```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone. Ellipses represent 95% confidence intervals around sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
# make the sample group look nice
nice <- function(s) {
  str_replace(s,"shallow","Shallow") %>%
    str_replace("main","(MHI)") %>%
    str_replace("nwhi","(NWHI)") %>%
    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
    map_chr(~str_c(.x,collapse=" "))
}

# get all the stuff we need to plot
pcap <- cap_depth$x %>%
  as_tibble(rownames="sample") %>%
  left_join(cd,by="sample") %>%
  mutate(
    # sample_group = case_when(
    #   island == "Oahu" & depth_zone == "shallow" ~ "O‘ahu (shallow)",
    #   .default = sample_group
    # ),
    sample_group = nice(sample_group),
    sample_group = fct_reorder(sample_group,depth)
  )
 centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))

# complement <- function(pal) {
#   s <- str_replace(pal,"^#","") %>%
#     str_match_all("..") %>%
#     map_chr(~{
#       ss <- .x[,1]
#       ss <- str_c("0x",ss) 
#       ss <- sprintf("%02x",255 - as.numeric(ss))
#       ss <- str_c("#",str_c(ss,collapse=""))
#     })
#   names(s) <- names(pal)
#   return(s)
# }

ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  # geom_point(color="black",shape=21,size=5)  +
  geom_point(color="black",size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
  # scale_color_manual(values = complement(sample_depth_pal)) + 
  # scale_color_manual(values=sample_depth_pal) + 
  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()
cat("{#fig:fig_cap}")
```

@@

@@

```{r rda-table,results="asis",dependson="rda-analysis"}
print_table(rda_table,"rda_table","Significant (marginal) terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
```

```{r alpha-box, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths). Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
library(patchwork)

```{r alpha-box2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
art <- shallow_alpha$richness_table %>% 
  mutate(island_group = str_to_sentence(island_group)) %>%
  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))

>>>>

---
title: "Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat"
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
        - nerr
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago. We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island. A total of 2284 individual crabs representing 90 morphospecies (19 families) were found. Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.  We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago. Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
bibliography: references/citations.bib
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: Table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

# Introduction {-}

The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet their nature and makeup remain poorly resolved. Coral reefs are among the most diverse habitats in the world, yet they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003]. In the face of these and other threats, understanding drivers of community structure is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019]. To date, the majority of efforts in this area have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018] or microbes [e.g., @desouzaImportanceDepthTemperature2023; @morrowEnvironmentalDriversMicrobial2016]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [[cite?]]{.mark}. Here, we investigate the factors influencing the community structure of a group of cryptic coral reef taxa across the Hawaiian Archipelago. <!--Say something about depth here, but I'm not sure what or how-->

Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974]. Such studies point to a combination of physical and biological factors<!--, including phyiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities. Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999]-->, but (with exceptions) a generalization has emerged <!--for the intertidal -->that communities' lower bathymetric limits are regulated by ecological interactions whereas upper limits are determined by abiotic factors. However, these models describe zonation at individual localities rather than community variation over larger spatial scales and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017]. In any case, it is likely that while ecological interactions are important for structuring communities at the local scale, external factors may exert stronger influence across greater distances [[cite?]]{.mark}.

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which consist of the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), which are the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities. Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Specifically, we used ARMS to sample brachyuran crabs, a diverse group of arthropods that comprise a considerable proportion of coral reef cryptofauna [289 species occur in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers. Additionally, given the strong reported differences between deep and shallow reefs across taxa [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation.

@@

A<!--nother--> likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->

Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.    Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI). We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts). We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.  Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.

@@

```{r fig-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

crab_sites <- cc$crabs_untransformed %>%
  sample_tibble() %>%
  distinct(hawaiian_name,island,island_group,lat,lon,.keep_all = FALSE) %>%
  mutate(
    island = case_when(
      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
      TRUE ~ as.character(hawaiian_name)
    ),
    island = fct_reorder(island,-lat)
  ) %>%
  st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%

@@

shift_box <- function(b) {
  if (b['xmin'] < 0) 
    b['xmin'] <- b['xmin'] + 360
  if (b['xmax'] < 0) 
    b['xmax'] <- b['xmax'] + 360
  return(b)
}

pmnm <- st_read(here("data","gis","hi_noaa_nwhi_papahanaumokuakea.shp"),quiet = TRUE) %>%
  st_shift_longitude()

pbox <- st_bbox(pmnm)
cbox <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")) %>% st_shift_longitude())
hbox <- hawaii %>%
  st_shift_longitude() %>%
  st_bbox()

box_before <- c(
  xmin = min(c(pbox['xmin'],hbox['xmin'],hbox['xmin'])),
  ymin = min(c(pbox['ymin'],cbox['ymin'],hbox['ymin'])),
  xmax = max(c(pbox['xmax'],cbox['xmax'],hbox['xmax'])) + 1.5,
  ymax = max(c(pbox['ymax'],cbox['ymax'],hbox['ymax']))
)

@@

pmnm <- pmnm %>%
  st_transform(transformer)

ggplot() +
  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
  geom_sf(data=pmnm,color=NA,fill="#6ed8f6",alpha=0.2) +
  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=island),size=3,shape=21,color="black") +
  scale_fill_manual(values=island_pal,name="Island") +
  coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
  scale_x_continuous(breaks = -seq(180, 155, -5)) +
  theme_minimal() +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
cat("{#fig:fig_map}")
```

@@

Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]). `r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated. The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).

```{r fig-rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

@@

```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone. Ellipses represent 95% confidence intervals around sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
# make the sample group look nice
nice <- function(s) {
  str_replace(s,"shallow","Shallow") %>%
    str_replace("main","(MHI)") %>%
    str_replace("nwhi","(NWHI)") %>%
    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
    map_chr(~str_c(.x,collapse=" "))
}

# get all the stuff we need to plot
pcap <- cap_depth$x %>%
  as_tibble(rownames="sample") %>%
  left_join(cd,by="sample") %>%
  mutate(
    sample_group = nice(sample_group),
    sample_group = fct_reorder(sample_group,depth)
  )
 centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))

ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  geom_point(color="black",shape=21,size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()
cat("{#fig:fig_cap}")
```

@@

```{r fig-alpha, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths). Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
library(patchwork)

```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
art <- shallow_alpha$richness_table %>% 
  mutate(island_group = str_to_sentence(island_group)) %>%
  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))

```{r fig-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site. Ellipses represent 95% confidence intervals around island groups', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
cds <- cc$crab_data_shallow

pcdata <- cds %>%
  select(where(is.numeric))

pca <- prcomp(pcdata)
explained <- (pca$sdev^2)/sum(pca$sdev^2)

bp <- cds %>%
  bind_cols(scores(pca))

loadings <- pca$rotation %>%
  as_tibble(rownames = "label") %>%
  mutate(
    PC1 = PC1*3,
    PC2 = PC2*3,
    hjust = if_else(PC1 < 0,1,0),
    xnudge = if_else(PC1 < 0,-0.02,0.02)
  )

ggplot(bp,aes(x=PC1,y=PC2)) + 
  geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color="black",alpha=0.1,show.legend = FALSE, size=0.07) +
  # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
  scale_fill_manual(values=island_group_pal) + 
  geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) + 
  geom_label(data=loadings,aes(x=PC1+xnudge,y=PC2,label=label,hjust=hjust)) + 
  xlab(str_glue("PC1 [{pc(explained[1])}]")) + 
  ylab(str_glue("PC2 [{pc(explained[2])}]"))
cat("{#fig:fig_pca}")
```

@@

```{r rda-table,results="asis",dependson="rda-analysis"}
print_table(rda_table,"rda_table","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
```
